---
title: "DATA624 - Project 2"
author: "Group 1"
date: "2025-07-19"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(VIM)
library(naniar)
library(DataExplorer)
library(caret)
library(tidymodels)
library(e1071)
library(reshape2)
library(lares)
```

## Data
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Download & load Data Dictionary
download.file(
  "https://github.com/hbedros/DATA624-Project2/raw/main/data/Data%20Columns%2C%20Types.xlsx",
  destfile = "data_dict.xlsx", mode = "wb"
)
data_dict <- read_excel("data_dict.xlsx")

# Training Data
download.file(
  "https://raw.githubusercontent.com/lalaexplore/DATA-624/main/Project%202/studentdata_model.xlsx",
  destfile = "train.xlsx", mode = "wb"
)
train_data <- read_excel("train.xlsx")

# Scoring Data
download.file(
  "https://github.com/hbedros/DATA624-Project2/raw/main/data/StudentEvaluation-%20TO%20PREDICT.xlsx",
  destfile = "score.xlsx", mode = "wb"
)
score_data <- read_excel("score.xlsx")

glimpse(train_data)
summary(train_data$PH)

```
## Initial EDA: Visualize the data
```{r}
summary(train_data)

plot_histogram(train_data)
plot_boxplot(train_data, by = 'PH')
plot_qq(train_data)
```

## Check missing values
```{r, echo=FALSE, message=FALSE, warning=FALSE}
vis_miss(train_data)
miss_var_summary(train_data)
gg_miss_var(train_data)

## Drop rows with NA
train_data <- train_data %>% filter(!is.na(PH))

summary(train_data$PH)
gg_miss_var(train_data)
```

## Impute missing predictors
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# kNN imputation on predictors (PH is not imputed!)
train_data$`Brand Code` <- as.factor(train_data$`Brand Code`)

train_data <- kNN(train_data, k = 5, imp_var = FALSE)

colSums(is.na(train_data))
gg_miss_var(train_data)
```

## EDA: Target Variable
```{r}
ggplot(train_data, aes(x = PH)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of PH", x = "PH", y = "Count")
```
- Most pH values are between ~8.4 and ~8.7.
- Distribution is fairly symmetric with a slight tail.

## Looking at predictors
#### Checking top variables correlated with PH
```{r, echo=FALSE, message=FALSE, warning=FALSE}
corr_cross(train_data, method = "spearman", max_pvalue = 0.05, top = 20, contains = 'PH')
```
- Top predictors: Mnf Flow, Bowl Setpoint, Usage cont, Filler Level.


## Top Predictor: Mnf Flow
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Find top correlated variable
numeric_cols <- train_data %>% select(where(is.numeric)) %>% select(-PH)

correlations <- cor(numeric_cols, train_data$PH, use = "complete.obs") %>%
  as.data.frame() %>%
  rownames_to_column("Variable") %>%
  arrange(desc(abs(V1)))

top_var <- correlations$Variable[1]

ggplot(train_data, aes(x = !!sym(top_var), y = PH)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_minimal() +
  labs(title = paste("PH vs", top_var), x = top_var, y = "PH")
```
- We observe extreme and clustered values for Mnf Flow, which we clean next.

### Clean Mnf Flow
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Clean extremes
train_data <- train_data %>%
  mutate(Mnf_Flow_Clean = ifelse(`Mnf Flow` < 0 | `Mnf Flow` > 250, NA, `Mnf Flow`))

# Impute cleaned values with median
median_flow <- median(train_data$Mnf_Flow_Clean, na.rm = TRUE)

train_data <- train_data %>%
  mutate(Mnf_Flow_Clean = ifelse(is.na(Mnf_Flow_Clean), median_flow, Mnf_Flow_Clean))

ggplot(train_data, aes(x = Mnf_Flow_Clean, y = PH)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  theme_minimal() +
  labs(title = "PH vs Cleaned Mnf Flow", x = "Cleaned Mnf Flow", y = "PH")
```

The cleaned data shows a weak positive relationship between Mnf Flow and pH, likely reflecting some operational influence. However, given the clustered, discrete flow settings and the weak effect, we explored the other top predictors, Bowl Setpoint, Usage cont, and Filler Level, to identify stronger relationships.

```{r}
top_vars <- c("Bowl Setpoint", "Usage cont", "Filler Level")

for (var in top_vars) {
  print(
    ggplot(train_data, aes(x = !!sym(var), y = PH)) +
      geom_point(alpha = 0.5) +
      geom_smooth(method = "lm", se = FALSE, color = "red") +
      theme_minimal() +
      labs(title = paste("PH vs", var), x = var, y = "PH")
  )
}

```

#### Other Top Predictors

- Bowl Setpoint: Weak positive relationship with pH - higher setpoints slightly increase pH, possibly by improving control or mixing.

- Usage cont: Weak negative relationship with pH - higher usage levels are associated with slightly lower pH, which may reflect depletion or wear effects.

- Filler Level: Weak positive relationship with pH - higher filler levels correspond to slightly higher pH, perhaps due to more stable process conditions.

No single variable among these showed a strong effect. This suggests that pH is influenced by a combination of factors, prompting further exploration of CO₂- and temperature-related variables.

## CO₂ & Temperature Variables

We examined variables logically tied to pH, including Carb Volume, Carb Pressure, Carb Flow, PSC CO₂, Carb Temp, and Temperature. Their Spearman correlations with pH were:
```{r}
co2_temp_vars <- c("Carb Volume", "Carb Pressure", "Carb Flow", "PSC CO2", 
                   "Carb Temp", "Temperature")

cor_df <- map_dfr(
  co2_temp_vars,
  ~ tibble(
      Variable = .x,
      Spearman = cor(train_data[[.x]], train_data$PH, method = "spearman", use = "complete.obs")
  )
) %>% arrange(desc(abs(Spearman)))

print(cor_df)
```

We also visualized the top three (Temperature, PSC CO₂, Carb Pressure):

- Temperature: Weak negative trend — higher temperatures slightly lower pH, consistent with CO₂ solubility decreasing at higher temperatures.

- PSC CO₂: Very weak negative trend — higher CO₂ levels lower pH, as dissolved CO₂ forms carbonic acid.

- Carb Pressure: Very weak positive trend — possibly due to operational settings or confounding factors.

```{r}
top_co2_temp <- cor_df$Variable[1:3]

for (var in top_co2_temp) {
  print(
    ggplot(train_data, aes(x = !!sym(var), y = PH)) +
      geom_point(alpha = 0.5) +
      geom_smooth(method = "lm", se = FALSE, color = "red") +
      theme_minimal() +
      labs(title = paste("PH vs", var), x = var, y = "PH")
  )
}

```

These CO₂ and temperature-related variables show weak but expected relationships with pH, supporting the hypothesis that pH depends on both gas solubility and process conditions. However, none of the variables individually explain much variance, suggesting that pH control is multivariate, and a predictive model that combines these factors may provide better insight and predictive power.